The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables

### 2014-12-01 draft.tex
CK edit final, plus some references

%\title{Modeling the Optical Spectra of Cataclysmic Variables: The Importance of Disk Winds}
%\title{Disk Winds Matter: Simulating the Optical Spectra of Cataclysmic Variables}
\title{The Impact of Disk Winds on the Optical Spectra of Cataclysmic Variables}
\author{J. H. Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}
%%\altaffiltext{1}{School of Physics \& Astronomy, University of Southampton,
  %%Southampton, SO17 1BJ, UK}

@@

\begin{abstract}
High-state non-magnetic cataclysmic variables (CVs) 
exhibit strong hydrogen and helium recombination lines in their optical 
spectra. These lines can be single peaked, and may be formed 
in the same bipolar outflow that is responsible for the
blue-shifted absorption lines seen in the ultraviolet (UV).
Here, we present results obtained by incorporating a `macro-atom' treatment into
our Monte Carlo radiative transfer code, which was
originally used to synthesize the UV spectra of disk-dominated, high-state CVs 
with a simple biconical wind. 
Using an identical model, we show that the same wind leaves a noticeable imprint 
on the optical spectrum, producing \ha\ and \heiiopt\ as well as
enhanced emission in the Ly-$\alpha$ and \heiiuv\ UV lines. 
% In addition, our improved treatment of recombination means that the bound-free continuum
% emission is significant, but does not yet overcome the intrinsic absorption edge
% present in the stellar atmosphere models used to treat the accretion disk.
We find that a boundary layer is not essential for the production
of He II recombination lines.
%  but also explore the effect of boundary layers
% with temperatures of $80-200$kK on the ionization state of the wind
% and resulting spectrum.
A short exploration of a 3-variable kinematic parameter space leads to models with higher densities 
that are able to produce single-peaked line emission, and also generate sufficient recombination emission
to fill in the Balmer jump photoabsorption edge intrinsic to our disk atmosphere spectra.
Finally, we present a synthetic spectrum of an improved optical model in 
and out of eclipse, which shows a striking similarity with 
the optical spectrum of the high inclination
Nova-like variable RW Tri. The densities near the disk plane in this model
are such that \civfull\ has a large collisionally excited component, in 
contrast to observations.
% We find that 
% a boundary layer is not essential to produce significant recombination to 
% He II, and if one is present we favour temperatures of $<10^5$K to avoid
% overionizing the UV resonance line species such as CIV. We also conduct Zanstra method
% calculations and estimate that one would infer a boundary layer temperature of around 80,000K
% from our benchmark calculation, despite it not being present in the model itself. This suggests that previous calculations
%done using this method are strongly dependent on both the assumed geometry and disk
%spectrum.
%We also note that
%the presence of a boundary layer becomes more noticeable in lower accretion rate systems.
\end{abstract}

@@

\section{Introduction}

% Outflows are ubiquitous  in accreting systems across many orders of magnitude in mass,
% from young stellar objects \citep{bunn1995} right up 
% to quasi-stellar objects \citep[QSOs;][]{ganguly2008}. Outflows can take the form of jets, such as those
% seen in X-ray binaries \citep{fender2006} or active galactic nuclei \citep[AGN;][]{marscher2006}, 
% but can also manifest as `winds' emanating from the accretion disk (REFS).
% Understanding the impact of accretion disk winds on spectral features has far-reaching implications. 
% They provide a feedback mechanism to limit the growth of black holes in AGN \citep{silkrees1998},
% and their ever-present nature implies a deep connection with the accretion process. This connection
% is important to understand in order to probe the universality of accretion \citep{mchardy2006, SKUK2012} 
% and to determine the true extent to which
% accretion disks conform to the `$\alpha$-disk' model proposed by 
% \cite{shakurasunyaev1973}. 
% Despite this, in many types of systems the impact of winds on observational
% features is essentially unknown, and many theoretical questions relating to their origin 
% and driving mechanism remain unanswered.

Cataclysmic variables (CVs) are systems in which a white dwarf accretes matter from a donor
star via Roche-lobe overflow. In non-magnetic, disk-dominated systems this accretion
is mediated by an accretion disk which forms around the white dwarf. 
Nova-like variables (NLs) are a particular subclass of objects in which the accretion disk
is in a permanent state of relatively high accretion rate 
($\dot{M} \sim 10^{-8}$~M$_{\odot}$~yr$^{-1}$),
making them the perfect laboratories to test the widely-used 
`$\alpha$-disk' model proposed by 
\cite{shakurasunyaev1973}.

For over three decades, it has been known that winds emanating from the accretion disk
are important in shaping the ultraviolet (UV) spectra of high-state CVs (Heap 1978), 
the most spectacular evidence being the P-Cygni like profiles of resonance lines such as 
\civfull\ (see e.g. Cordova \& Mason 1982\nocite{cordova1982}).
Despite this, in many types of systems the impact of winds on observational
features is essentially unknown, and many theoretical questions relating to their origin 
and driving mechanism remain unanswered.

While the effect of the wind on the UV is at least partially clear, it may also have more subtle 
effects across the spectral range. It has been proposed that an accretion disk wind
can provide a mechanism for single-peaked emission lines (Murray \& Chiang 1996), 
such as those seen in Nova-like variables and, especially, the SW Sex stars \citep{HSK86, DR95}.
It is particularly telling that single-peaked 
lines in high-inclination systems can be seen even in eclipse, suggesting 
that a spatially extensive wind may be the source of emission (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence of the influence of winds in the optical can occasionally 
be occasionally seen in P-Cygni like profiles observed in lines such as \ha\ 
and He \textsc{i} $\lambda5876$ \citep{RN98}.

% Somewhere, probably the introduction, we need to discuss alternatives for
% explaining the Balmer Jump problem. The other major possibility is that
% the disk atmospheres are not correct, see e.g Shaviv or that
% backscattering fromt he wind changes the disk structure. This is the main
% weakness in our model approach, because we cannot calculate the atmosphere
% ourselves. (Christian and I have had some discussions with Hillier, who is
% here in Aspen, and is the author of CMFGEN.  There was a paper on which he
% was a co-author where 1-d models of winds were summed to ³make² a CV wind.
% These used CMFGen and to calculate the structure of the wind, and so may
% include the backr]reaction of the wind on the disk photosphere.  He said
% he would check if these models had Balmer jumps, but you might want to
% find the paper).

It has also been proposed
that recombination continuum emission from a disk wind or `transition region'
could be responsible for filling in the Balmer absorption edge  \citep{KLWB98}. 
This provides a potential explanation for
the absence of a pronounced `Balmer jump' in optical spectra of CVs, 
which has never been adequately produced by using stellar atmospheres 
to model the accretion disk (see e.g. Haug 1987)\nocite{haug1987}. 
However, it is also possible that accretion disks do not possess
and intrinsic deep Balmer absorption edges.
This depends to the optical depth at the edge
and can change due to decrease due to less temperature variation with height
near the disk photosphere (see e.g. Shaviv et al. 1991; 
but also more recent work by Idan et al. 2010).
\nocite{idanshaviv2010} \nocite{shaviv1991}

% This dense disk atmosphere or wind transition region
% necessary to produce the required recombination emission
% could also explain the Fe~\textsc{ii} `curtain'
% seen in near-UV observations of the high inclination dwarf nova OY Car in quiescence

% In addition, \cite{KLWB98} suggested
% that recombination continuum emission from a disk wind or atmosphere
% could be responsible for filling in the Balmer absorption edge. This provides
% a potential explanation for
% the absence of a pronounced `Balmer jump' in optical spectra of CVs, which has never been adequately produced 
% by theoretical models or spectral synthesis.
% This dense disk atmosphere or wind transition region
% necessary to produce the required recombination emission
% could also explain the Fe~\textsc{ii} `curtain'
% seen in near-UV observations of the high inclination dwarf nova OY Car in quiescence 
% \citep{horne1994}.

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (Top; Beuermann et al. 1992),
IX Vel (Middle; Beuermann \& Thomas 1990)
and RW Tri in eclipse (Bottom; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively.
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially extensive disk wind, and there is even a suggestion
of a (potentially wind-formed) recombination continuum in the eclipsed spectrum.}
\label{novalikes}
\end{figure}

The first attempts to model disk winds in CVs applied techniques used in stellar
wind modeling to an accretion disk geometry (Mauche \& Raymond
1987; Drew 1987; Vitello \& Shlosman 1988). These efforts showed that mass loss rates of
$\sim10\%$ of the accretion rate
were required to produce the observed line strengths.
Using spectral synthesis codes to model resonance scattering,
Shlosman \& Vitello (1993; hereafter SV93)\nocite{SV93} and \cite{KWD95}
were both successful in reproducing the observed spectra such as the P-Cygni like
profiles in C\textsc{iv}~$\lambda1549$\AA.
Long \& Knigge (2002; hereafter LK02)\nocite{LK02} presented a code that
solved self-consistently for thermal and ionization balance in the wind
according to the methods of Mazzali \& Lucy (1993; ML93)\nocite{ML93}. They successfully reproduced
UV resonance lines profiles and presented a synthesized spectrum of Z Cam
that agreed well with the observed features. Noebauer et al. (2010) used the same code 
to carry out more detailed comparisons to RW Tri and UX UMa. In addition to these 
radiative transfer calculations, some work has also been done using hydrodynamic wind models, 
such as that by \cite{pkdh2002}, and more recently by \cite{kusterer2014} with application to
AM Canum Venaticorum stars.

Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
but incorporating additional `macro-atom' techniques suggested by Lucy (2002, 2003). 
Our aim is to model high-state CVs over a wide spectral range.
In particular, we hope to assess the potential contribution
to the optical spectrum from a disk wind model which preserves the UV features.
It will then be possible to see how closely
our models resemble the spectra in figure 1 as we change wind parameters.
We will investigate whether a wind can produce significant
recombination emission lines and whether the recombination
continuum emission is sufficient to mask the Balmer jump from a disk atmosphere.
We will also learn if the mechanism for single-peaked line emission
proposed by \cite{MC96} works with plausible model parameters.
In section 2 we present the code and provide a brief description of macro-atoms 
and the radiative transfer methods we employ. 
We describe the kinematics and geometry of our biconical wind model in section 3.
In section 4, we present a benchmark model, which is explored on in section 5 
with a small-scale parameter search. In section 6, we present a
model which better reproduces the optical spectrum of a NL, and, 
finally, in section 7, we summarise our findings.

% Here, we present work using the same Monte Carlo radiative transfer code as LK02, 
% \py, but incorporating additional 
% `macro-atom' techniques suggested by Lucy (2002, 2003). 
% This improves our treatment of the Hydrogen and Helium
% optical lines, as well as recombination continuum emission. 
% Our goal is to investigate the effect of existing wind models
% on the optical spectra of CVs.

% {\bf
% \noindent Fairly happy with this as an introduction. Could maybe
% discuss the boundary layer background here instead of in section 2.
% }

@@

\section{Radiative Transfer}

\py is a Monte Carlo (MC) radiative transfer code which uses
the Sobolev approximation to treat line transfer. It operates 
in two distinct cycles. First, the ionization state, level populations
and temperature structure are calculated. This is done by
flying MC energy quanta (`photons') through a pre-defined wind geometry and recording
the relevant estimators, before choosing  a new electron temperature 
in an attempt to balance heating and cooling in the plasma.
This process is repeated until the code converges on a 
solution. Second, the spectrum is synthesized by tracking
the energy packets through the wind and computing the 
emergent spectra at a number of user-specified viewing angles.

The code is capable of operating in a number of different
regimes, both in terms of the scale of the system, and the 
shape and characteristic energies of the illuminating spectrum. 
It was originally developed by LK02 in order to model the UV spectra
of CVs with a simple biconical disk wind model.. 
Higginbottom et al.\nocite{higginbottom2013} (2013; hereafter H13) used the same code to
present a benchmark model for broad absorption line (BAL) QSOs.
As part of this effort, the ionization scheme was improved so that 
the code is now capable of treating arbritrary SEDs.
%the power-law X-ray spectrum exhibited by AGN.
% This involved using a spectral model the mean intensity in a cell
% computed from MC estimators for, instead of using a dilute blackbody
% approximation. Compton processes
% were also incorporated into the heating and cooling balance.
In addition, Sim et al. (2005; hereafter SDL05)\nocite{simmacro2005} modeled
Brackett and Pfund line profiles in young-stellar objects (YSOs) using \textsc{Python,} which included
implementing a `macro-atom' mode in order to correctly treat recombination lines.

\subsection{Macro-atoms}

Previously to SDL05, \py used a two-level atom approximation \cite[see e.g.][]{mihalas}. This approximation is 
fast, and works well as a first approximation when treating so-called `resonance lines' (such as \civfull) 
in which the excited electron is strongly coupled to the ground state.
However, in a number of situations, for example, when a level is primarily populated from above, then
this becomes a poor representation. 
To reproduce recombination lines and lines in which the lower level is not the ground state, 
it is therefore important to use an improved line transfer technique and level population solver.

By quantising matter into `macro-atoms', and radiant and kinetic 
energy into indivisible energy packets (r- and k- packets respectively), 
Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) showed that it is possible 
to `asymptotically reproduce the emissivity of a gas in statistical equilibrium 
without simplifying the treatment of line transfer'. 
His macro-atom scheme allows all possible transition paths from a given level and a full non-Local 
Thermodynamic Equilibrium (NLTE) solution for the level populations based on MC estimators. 
For a full description consult L02 and L03. 
Macro-atoms have been used to model astrophysical plasmas in 
Wolf-Rayet star winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kasen2006, kerzendorfsim} and YSOs (SDL05).
% Recombination is expected to be an important process in the optical
% spectra of CVs, and they thus make a perfect application for Lucy's 
% advances in MC methods.

\subsection{Ionization and Excitation}

Whereas SDL05 used \py itself to apply the macro-atom 
technique to a Hydrogen-only model, we
adopt a hybrid ionization and excitation scheme. 
Species can be treated as either full macro-atoms
or using the two-level atom approximation.
In the latter case, they still follow the indivisible packet constraint.
This allows one to preserve the fast treatment of resonance lines while providing 
an improved representation of processes with more complex
transition paths such as recombination. This is important when modeling features 
such as the Balmer lines and also recombination {\sl continuum} emission.

% In our hybrid ionization and excitation scheme, 
% simple-ions have their ionization fractions
% calculated using a modified Saha equation, and internal level populations are
% calculated from LTE. For macro-atoms, we solve the equations of statistical 
% equilibrium using MC estimators for the radiation field. This enables us to 
% determine the ionization and excitation state of the ions in question.

\subsubsection{Macro-atoms}
% In order to correctly model lines in which the lower level is not the ground state 
% (such as H$\alpha$ \& H$\beta$), and improve the treatment of recombination,
% we treat H \& He as macro-atoms. This is because 
% the levels responsible for the line emission are populated mostly from above.
% Therefore, 
We treat H and He as macro-atoms. 
Macro-atom NLTE level populations and ion fractions are calculated by solving 
the statistical equilbrium equations between each pair of levels. The bound-bound 
de-excitation rate, ${\cal R}_{ul}$, in an ion is given by

\begin{equation}
{\cal R}_{ul} = \beta_{lu} A_{ul} n_u + \beta_{lu} B_{ul} n_u J_{lu} + C_{ul} n_u n_e,\label{eq:nlte_rul}
\end{equation}

where $u$ and $l$ denote the upper and lower levels, $C$ is the collisional rate coefficient
and $A$ and $B$ are the Einstein coefficients. $\beta_{lu}$ is the probability that a 
given line photon will escape the Sobolev region. $J$ is the MC estimator for the mean intensity in the line as 
recorded during the photon propagation. The corresponding excitation rate is then

\begin{equation}
{\cal R}_{lu} = \beta_{lu} B_{lu} n_l J_{lu} + C_{lu} n_l n_e.
\end{equation}

We also calculate the photoionization rates between a lower level $l$,
and the continuum state (or, in the case of ions with more than one 
bound electron, the ground state of the upper ion), $\kappa$,

\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu} \sigma_{\nu}}{h \nu} d\nu,
\end{equation}

where $\sigma_{\nu}$ is the photoionization cross section and $J_{\nu}$ is the mean intensity. 
The recombination rate is given by

\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e, \\
\end{equation}

where $\alpha_{\kappa l}$ is the recombination
coefficient to level $l$. 
This treatment means that radiative and collisional rates to and from all 
levels are considered when calculating both the ionization state, and the level populations, 
although we neglect photoionization to excited levels of the upper ion. 
The \cite{vanregemorter} approximation is used for collisional transitions,
and thus we assume collisions between radiatively forbidden transitions are
not important in determining level populations.
This is a reasonable approximation in the regimes considered here.
%{2002A&A...384..725L}.  source for beta sobolevs
%$\beta_{lu}=\frac{1}{\tau_{lu}}[1-\exp(-\tau_{lu})]$

\subsubsection{Two-level atoms}

Lines from all other elements are treated with the two-level atom approximation. 
This is a first order approximation, which is appropriate 
when treating resonance lines in which the lower level is at or close to the 
ground state. 
For these elements, we follow LK02 in using the ionization scheme presented by Mazzali \& Lucy (1993). 
Specifically, we use the formula
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R} \label{ionization}
\end{equation}
to solve for the relative ionization fractions. In this equation, the `starred' term on 
the right represents the Saha abundances, in accordance with the notation of Mihalas (1982). 
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. LK02 provide a full description of our implementation.
This ionization scheme is appropriate here because
the SED can be approximated well by a dilute blackbody due to the 
absence of a power-law X-ray source, unlike with AGN (see e.g. H13).
For two-level atoms, we assume LTE level populations.
% This ionization scheme is appropriate here because, unlike with AGN,
% the SED can be approximated well by a dilute blackbody due to the 
% absence of a power-law X-ray source, and therefore the above modification to the Saha equation
% is appropriate; Second, this provides continuity from LK02 and allows us to do a like-for-like comparison.
% This is especially useful in assessing the improvement in Hydrogen line emission, and any changes
% that might occur in the ionization stages of the resonance line species such as CIV. 
% Excitation is then calculated in LTE. 
% This is because the 
% level populations are only important in working out the resonance scattering 
% opacity through the wind when modeling resonance lines.

>>>>

\title{The Impact of Accretion Disk Winds on the Optical Spectra of Cataclysmic Variables}
\author{J. H. Matthews, C. Knigge, K. S. Long, S. A. Sim \& N. Higginbottom}

@@

\begin{abstract}
Many high-state non-magnetic cataclysmic variables (CVs) exhibit
blue-shifted absorption or P-Cygni profiles in ultraviolet
(UV) resonance lines, such as C~{\sc iv}~1550\AA. These features
imply the existence of powerful accretion disk winds in
CVs, and considerable effort has gone into modelling UV line formation
in such outflows. By contrast, much less attention has been paid to
the potential impact of these disk winds on the {\em optical} spectra
of CVs. Here, we use an improved version of our Monte Carlo ionization
and radiative transfer code to investigate whether disk wind models that
produce realistic UV line profiles are also likely to generate
observationally significant recombination line and continuum emission
in the optical waveband. In particular, we test whether outflows may 
be responsible for the single-peaked emission line profiles often seen
in high-state CVs and for the weakness of the Balmer absorption edge
(relative to simple models of optically thick accretion disks).
We find that a standard disk wind model that is successful in
reproducing the UV spectra of CVs does leave a noticeable imprint on
the optical spectrum as well, particularly for systems viewed at high
inclination. The strongest optical wind-formed recombinations lines are \ha\ and
\heiiopt\, although the outflow also produces emission in the Ly-$\alpha$
and \heiiuv\ UV lines. However, all of the optical lines produced by
this default wind model are double-peaked, and the effect of the wind
on the Balmer jump is modest. We carry out a limited exploration of
the disk wind parameter space, focusing on those parameters that affect the
density near the base of the wind. This shows that higher-density
outflows can produce both single-peaked emission lines and a
recombination continuum sufficient to fill in the Balmer
jump, although such models also tend to predict stronger-than-observed 
collisionally excited UV emission lines. Our main conclusion is that
disk wind emission is likely to have a significant impact on the
optical spectra of high-state CVs. 
\end{abstract}

@@

\section{Introduction} 
\label{sec:intro}

Cataclysmic variables (CVs) are systems in which a white dwarf
accretes matter from a donor star via Roche-lobe overflow. In
non-magnetic systems this accretion is mediated by a Keplerian disk
around the white dwarf (WD). Nova-like variables (NLs) are a subclass
of CVs in which the accretion disk is always in a relatively
high accretion rate state ($\dot{M} \sim
10^{-8}$~M$_{\odot}$~yr$^{-1}$).  This makes NLs an excellent
laboratory for studying the properties of steady-state accretion
disks.

It has been known for a long time that winds emanating from the
accretion disk are important in shaping the ultraviolet (UV) spectra
of high-state CVs (Heap 1978). The most spectacular evidence for such
outflows are the P-Cygni-like profiles seen in UV resonance lines such as
\civfull\ (see e.g. Cordova \& Mason
1982\nocite{cordova1982}). Considerable effort has been spent over the
years on understanding and modelling these UV features (e.g. Drew \&
Verbunt 1985; Mauche \& Raymond
1987; Drew 1987; Shlosman \& Vitello 1993; [hereafter
SV93]\nocite{SV93}; Knigge, Woods \& Drew 1995 [hereafter
KWD95]\nocite{KWD95}; Knigge \& Drew 1997; Knigge et al. 1997; Long \&
Knigge 2002 [hereafter LK02]\nocite{LK02}, Noebauer et al. 2010;
Puebla et al. 2011). The basic picture emerging from these efforts is
of a slowly accelerating, moderately collimated bipolar
outflow that carries away $\simeq 1\% - 10\%$ of the accreting
material. State-of-art-simulations of line formation in this type
of disk wind can produce UV line profiles that are remarkably similar
to observations.

Much less is known about the effect of these outflows on the optical
spectra of high-state CVs. These spectra are typically characterized
by H and He emissions superposed on a blue continuum. In many
cases, and particularly in the SW~Sex subclass of NLs
\citep{HSK86,DR95}, these lines are single-peaked. This is contrary to
theoretical expectations for lines formed in accretion disks, which
are predicted to be double-peaked \citep{smak1981, hornemarsh1986}. 
{\em Low-state} CVs (dwarf novae in quiescence) do, in fact,
exhibit such double-peaked lines \citep{marshhorne1990}(MOREREF). 
{\bf JM: Other things to consider; transition of line profiles
from quiescence to outburst in DNe, emission lines in absorption cores
trends of lines and continuum with inclination, flat
Balmer decrement}.

Murray \& Chiang (1996) have shown that the presence of disk winds may
offer a natural explanation for the single-peaked optical emission lines in
high-state CVs, since they can strongly affect the radiative transfer
of line photons. Strong support for a significant wind contribution to the
optical emission lines comes from observations of eclipsing
systems. There, the single-peaked lines are often only weakly
eclipsed, and a significant fraction of the line flux remains visible
even near mid-eclipse (REF). This points to line formation in a spatially
extended region, such as a disk wind (see figure~\ref{novalikes}).
%%(Honeycutt, Schlegel, \& Kaitchuck 1986; Dhillon \& Rutten 1995). 
Further evidence for a wind contribution to the optical lines comes
from isolated observations of P-Cygni-like line profiles even in optical
lines, such as \ha\ and He \textsc{i} $\lambda5876$ \citep{RN98}.
{\bf JM: need to mention this as pointed out by CK.}

Could disk winds also have an impact on the UV/optical {\em continuum}
of high-state CVs? This continuum is usually thought to be dominated
by the accretion disk and modelled by splitting the disk into
a set of concentric, optically thick, non-interacting annuli following
the standard $T_{eff}(R) \propto R^{-3/4}$ radial temperature
distribution \citep{shakurasunyaev1973}. In such
models, each annulus is taken to emit either as a blackbody or,
perhaps more realistically, as a stellar/disk atmosphere model
(REF). In the latter case, the local surface gravity, $\log{g}(R)$, is
assumed to be set solely by the accreting WD, since self-gravity is
negligible in CV disks.

Attempts to fit the observed spectral energy distributions (SEDs) of
high-state CVs with such models have met with mixed success. In
particular, the SEDs predicted by most stellar/disk atmosphere models 
are too blue in the UV (REFs) and exhibit
stronger-than-observed Balmer jumps in absorption (REFs). One possible
explanation for these problems is that these models fail to capture
all of the relevant physics. Indeed, it has been argued that a
self-consistent treatment can 
produce better agreement with observational data (e.g. Shaviv et
al. 1991;  but also more recent work by Idan et al. 2010). 
\nocite{idanshaviv2010} \nocite{shaviv1991}. However, an alternative 
explanation, suggested by \cite{KLWB98}, is that recombination continuum
emission from the base of the disk wind might fill in the disk's
Balmer absorption edge and flatten the UV spectrum.

\begin{figure}	%fullpage
\centering
\includegraphics[width=0.8\textwidth]{figures/fig1.eps}
\caption{Optical spectra of three nova-like variables: 
RW Sex (Top; Beuermann et al. 1992),
IX Vel (Middle; Beuermann \& Thomas 1990) %% CITE THESE
and RW Tri in eclipse (Bottom; Groot et al. 2004).
The data have been digitized from the respective publications and a few of the more
prominent emission and absorption lines are marked. 
These systems have inclinations of $30^\circ$, $60^\circ$ and $70^\circ$ 
respectively.
The trend of increasing Balmer line emission with inclination can be seen.
In RW Tri strong single-peaked emission in the Balmer lines is seen even
in eclipse, indicating that the lines may be formed in a spatially
extensive disk wind, and there is even a suggestion 
of a (potentially wind-formed) recombination continuum in the eclipsed
spectrum.}
\label{novalikes}
\end{figure}

%% CITATIONS FROM CAPTION
\nocite{groot2004}

Here, we carry out Monte Carlo radiative transfer simulations in
order to assess the likely impact of accretion disk winds on the
optical spectra of high-state CVs. More specifically, our goal is to
test whether disk winds of the type developed to account for the UV
resonance lines would also naturally produce significant amounts of  
optical line and/or continuum emission. In order to achieve this, we
have implemented the `macro-atom' approach developed by Lucy
(2002, 2003) into the Monte Carlo ionization and radiative transfer
code described by LK02 (a process initiated by SDL05). With this
upgrade, the code is able to deal correctly with processes involving
excited levels, such as the recombination emission produced by CV
winds.

The remainder of this paper is organized as follows. In Section~2 we
briefly describe the code and and the newly implemented macro-atom
approach. In Section~3, we describe the kinematics and geometry of our
disk wind model. 
In Section~4, we present spectra simulated from the benchmark model
employed by LK02, and in Section~5, we carry out a limited sensitivity
study to explore how the predicted spectra change as a function of
some key parameters in our model. In Section~6, we summarize our
findings.

@@

\section{\sc{python}: A Monte Carlo Ionization and Radiative Transfer Code}

\py is a Monte Carlo (MC) ionization and radiative transfer code which
uses the Sobolev approximation to treat line transfer. It has already
been described extensively in LK02, SDL05 and H13, so here we provide
only a brief summary of its operation, focusing particularly on new
aspects of our implementation of macro-atoms into the code.

\subsection{Basics}

\py operates in two distinct stages. First, the ionization state,
level populations and temperature structure are calculated. This is
done iteratively, by 
propagating several populations of MC energy quanta (`photons')
through a model wind. The geometric and kinemetic properties of the
outflow are specified on a pre-defined spatial grid. In each of these
iterations (``ionization cycles''), the code records estimators that 
characterize the radiation field in each grid cell. At the end 
of each ionization cycles, a new electron temperature is calculated
that more closely balances heating and cooling in the 
plasma. The radiative estimators and updated electron
temperature are then used to revise the ionization state of the wind,
and a new ionization cycle is started. The process is repeated until
heating and cooling are balanced throughout the wind.

This converged model is used as the basis for the second set of
iterations (``spectral cycles''). In these, the emergent spectrum over
the desired spectral range is synthesized by tracking populations of
energy packets through the wind and computing the emergent spectra at
a number of user-specified viewing angles.

\py is designed to operate in a number of different
regimes, both in terms of the scale of the system, and in terms of the
characteristics of the underlying radiation field.
It was originally developed by LK02 in order to model the UV spectra
of CVs with a simple biconical disk wind model. Sim et al. (2005;
hereafter SDL05)\nocite{simmacro2005} used the code to model Brackett
and Pfund line profiles of H in young-stellar objects (YSOs). As part
of this effort, they implemented a `macro-atom' mode (see below) in
order to correctly treat hydrogen recombination lines with
\py. Finally, Higginbottom et al.\nocite{higginbottom2013} (2013;
hereafter H13) used \py to model broad absorption line (BAL) QSOs. For
this application, an improved treatment of ionization was implemented,
so that the code is now capable of dealing with arbritrary
photo-ionizing SEDs, including non-thermal and multi-component ones.

\subsection{Ionization and Excitation: ``Simple Atoms''}

Prior to SDL05, the relative ionization fractions for all atomic
species were estimated via the modified Saha equation (Mazzali \&
Lucy 1993)  
\begin{equation}
\frac{n_{j+1} n_e}{n_j} = W [\xi + W(1-\xi)]
\left(\frac{T_e}{T_R}\right)^{1/2}
\left(\frac{n_{j+1}n_e}{n_j}\right)^*_{T_R}, \label{ionization}
\end{equation}
Here, the `starred' term on the right represents the Saha (LTE)
abundances, in accordance with the notation of Mihalas (1982).  
$W$ is an effective dilution factor, $\xi$ is the
fraction of recombinations going directly to the ground state, and
$T_R$ and $T_e$ are the radiation and electron temperature
respectively. This simple ionization scheme produces reasonable
results when the photoionizing SED can be approximated by a dilute
blackbody. This is the case for high-state CVs. (As noted above, an
improved, but more complex treatment of ionization that is appropriate
for more complex SEDs is described in H13.)

Similarly, the relative excitation fractions within each ionization
stage of a given species were estimated via a modified Boltzmann
equation,
\begin{equation}
\frac{n_{jk}}{n_j} = W g_k \exp{-E_k/kT_R} / z_j(T_R),
\end{equation}
where $n_{jk}$ is the population of level $k$ in ionic stage $j$,
$E_k$ is the energy difference between level $k$ and the ground state,
and $z_j(T_R)$ is the partition function of ionic stage $j$.

Finally, \py modelled all bound-bound processes as transitions
within a simple two-level atom \cite[see e.g.][]{mihalas}. This
approach works reasonable well for resonance  
lines, such as \civfull, in which the lower level is the ground state.  
However, it is a poor approximation for many other
transitions, particularly those where the upper level is 
is primarily populated from above. Thus an improved method for
estimating excited level populations and simulating line transfer is
needed in order to model recombination lines and continua.

\subsection{Ionization and Excitation: Macro-Atoms}

Lucy (2002, 2003\nocite{lucy2002, lucy2003}; hereafter L02, L03) 
has shown that it is possible to calculate the emissivity of a gas in
statistical equilibrium accurately by quantising matter into
`macro-atoms', and radiant and kinetic energy into indivisible energy
packets (r- and k- packets, respectively). His macro-atom scheme
allows for all possible transition paths from a given level and
provides a full non-local thermodynamic equilibrium (NLTE) solution
for the level populations based on MC estimators. The macro-atom
technique has already been used to model Wolf-Rayet star
winds \citep{sim2004}, AGN disk winds \citep{simlong2008, tatum2012},
supernovae \citep{kasen2006, kerzendorfsim} and YSOs (SDL05). A full 
description of the can be found in L02 and L03.

Briefly, macro-atom NLTE level populations and ionization fractions
are calculated by solving the statistical equilbrium equations between
each pair of levels. For example, the bound-bound de-excitation rate,
${\cal R}_{ul}$, in an ion is given by  
\begin{equation}
{\cal R}_{ul} = \beta_{lu} A_{ul} n_u + \beta_{lu} B_{ul} n_u J_{lu} +
C_{ul} n_u n_e,\label{eq:nlte_rul},
\end{equation}
where $u$ and $l$ denote the upper and lower levels, $C$ represents the
collisional rate coefficients and $A$ and $B$ represent the Einstein
coefficients. The quantity $\beta_{lu}$ is the probability that a
given line photon will escape the Sobolev region. Finally, $J$ is the
MC estimator for the mean intensity in the line, as recorded during
the photon propagation. The corresponding excitation rate is then 
\begin{equation}
{\cal R}_{lu} = \beta_{lu} B_{lu} n_l J_{lu} + C_{lu} n_l n_e,
\end{equation}
where we implicitly include a correction for stimulated emission in
our definition of the estimate for $J_{ul}$.

In our implementation of the macro-atom approach, we also explicitly
take into account the photoionization and collisional ionization rates
between a lower level, $l$, and the continuum (or, in the case of ions
with more than one bound electron, the ground state of the upper ion),
$\kappa$,
\begin{equation}
{\cal R}_{l \kappa}= n_l \int_{\nu_0}^{\infty} \frac{ 4 \pi J_{\nu}
  \sigma_{\nu}}{h \nu} d\nu + C_{l \kappa} n_l n_e.
\end{equation}

Here, $\sigma_{\nu}$ is the photoionization cross section, and $J_{\nu}$
is the mean intensity. The corresponding recombination rate is given
by 
\begin{equation}
{\cal R}_{\kappa l} = \alpha_{\kappa l} n_{\kappa} n_e + C_{\kappa l}
n_\kappa n_e, \\
\end{equation}
where $\alpha_{\kappa l}$ is the radiative recombination coefficient
to level $l$. This treatment means that radiative and collisional
rates to and from all levels are considered when calculating both the
ionization state and the level populations, although we neglect 
ionization directly to excited levels of the upper ion. The
\cite{vanregemorter} approximation is used for collisional
transitions, and thus we assume that collisions between radiatively
forbidden transitions are not important in determining level
populations. This is a reasonable approximation in the regimes
considered here.

\subsection{Ionization and Excitation: A Hybrid Approach}

SDL05 implemented a macro-atom treatment of Hydrogen in \py and used
this to predict the observable properties of a pure Hydrogen wind
model for YSOs. Our goal here is to simultaneously model the optical
and ultraviolet spectra of high-state CVs. Since the optical spectra
are dominated by H and He recombination lines, both of these species
need to be treated as macro-atoms. The UV spectra, on the other hand,
are dominated by resonance lines associated with metals. This means we
need to include these species in our models, but they can be treated 
with our (much faster) simple-atom approach. We have therefore
implemented a hybrid ionization and excitation scheme into \py. Any
atomic species can now be treated either in our simple-atom
approximation or with the full macro-atom machinery. For our CV
models, we treat H and He as macro-atoms and all metals as
simple-atoms.

\subsection{Atomic Data}
